Forcing for Isabelle

### 2018-12-28 LICS/conclusion.tex
empezando a escribir la conclu

The proof of the Powerset Axiom does not present difficulties.

By G\"odel's Second Incompleteness Theorem we cannot  prove that
there exists a model of $\ZF$  (assuming that our base theory is not
stronger than $\ZF$ to start with), but for consistency proofs,
usually a model of only a finite subset of $\ZF$ suffices. Recall that
an equivalent formulation of the Foundation Axiom states that the 
universe of sets can be decomposed in a transfinite hierarchy of
sets. 
\begin{theorem}
  Let $V_{\al}\defi\union\{\P(V_\be) : \be<\al\}$ for each ordinal
  $\al$. Then each $V_\al$ is a set and 
  $\forall x. \exists\al .\ \Ord(\al) \y x\in V_\al$.  
\end{theorem}
%
Then the following meta-theoretic result by Montague applies:
%
\begin{theorem}[Reflection Principle]\label{th:reflection-principle}
  For every finite $\Phi\sbq\ZF$, $\ZF$ proves: ``There exists
    unboundedly many $\al$ such that $V_\al\models \Phi$.''
\end{theorem}
%
Since the sets $V_\al$ are well founded, we can repeat the
proof of Lemma~\ref{lem:wf-model-implies-ctm} to obtain a ctm of
$\Phi$.

The proof of the Replacement Axiom, both in Kunen and in
other presentations \cite{neeman-course} require a relativized proof
of the Reflection Principle.

>>>>

The proof that the axioms of $\ZF$ hold in generic extensions is 
the first important landmark in the path of formalization of the
theory of forcing. The task of proving the Fundamental Theorems can be
considered a part of that landmark, but from a workflow point of view
this can be done independently, and one of our design decisions was to
postpone that task.  Next, various forcing notions must be developed
in order to obtain extensions satisfying diverse properties. This
yields relative consistency results.

We first discuss the axioms we are now missing. The proof of the
Powerset Axiom does not present difficulties, and it is one of the
first goals in the immediate future. This is also the case with the
Axiom of Infinity; it can be proved in a direct but cumbersome way. We
preferred to wait to have a full-fledged inteface between ctms and the
locales providing recursive constructions from Paulson's
\isatt{Constructibility} session. Then the current proof will hold
with no extra assumptions. The same goes for the results $M\sbq M[G]$
and $G\in M[G]$.

The Replacement Axiom, however, requires some more work to be
done. Both in Kunen and in other presentations \cite{neeman-course}
it requires a relativized version (i.e., showing that it holds for $M$) of
the \emph{Reflection Principle}. In order to state this meta-theoretic
result by Montague, recall that
an equivalent formulation of the Foundation Axiom states that the 
universe of sets can be decomposed in a transfinite hierarchy of
sets. 
\begin{theorem}
  Let $V_{\al}\defi\union\{\P(V_\be) : \be<\al\}$ for each ordinal
  $\al$. Then each $V_\al$ is a set and 
  $\forall x. \exists\al .\ \Ord(\al) \y x\in V_\al$.  
\end{theorem}
\begin{theorem}[Reflection Principle]\label{th:reflection-principle}
  For every finite $\Phi\sbq\ZF$, $\ZF$ proves: ``There exists
  unboundedly many $\al$ such that $V_\al\models \Phi$.''
\end{theorem}

This result has also other applications; in particular, it dispenses
the need of ctms of the whole of $\ZF$ for consistency proofs (by
repeating  the proof of Lemma~\ref{lem:wf-model-implies-ctm}, now
starting with a  $V_\al$ satisfying $\Phi$).
